Draft

### 2022-06-20 Final/FinalReport.qmd
fixed latex

Abortion is one such divisive issue in the United States. Both pro-choice and pro-life groups have a history of mobilizing in states across the country in support of legislation for their respective sides (Ziegler, 2020). However, following the Supreme Court's 1973 decision in Roe v. Wade, a ruling which protected an individual’s right to have an abortion before fetal viability, the issue has risen in political salience. Both the pro-choice and pro-life movements have gained national prominence, and the two major political parties have polarized around the issue, with the Democratic Party in favor of and the Republican Party against policies legalizing and increasing access to abortion (Weinberger 2022). Abortion has also increasingly become a key issue that voters consider when making their choice at the ballot box, with an increasing share of Americans identifying as "single-issue voters" regarding abortion (Brenan 2020).

@@

The dataset used in this analysis observes attitudes on the justifiability of abortion among respondents in in the United States across six “waves” of the World Values Survey (1982-2011), which is administered every few years and collects information about people’s values and beliefs worldwide, alongside basic demographic characteristics. The data are collected via face-to-face interviews at the respondents’ homes. In this data set, the response variable, justifiability of abortion, is a numerical measure on a scale of 1 to 10 on the individual person’s attitude toward whether abortion is justifiable or not.  Individuals responded 1 for “abortion is never justified” and 10 for “abortion is always justified.

Our research question is as follows: “Do an individual’s political ideology and their personal attitudes/preferences towards other issues, such as the importance of religion in their life and their respect for authority, among others, predict their attitude on the justifiability of abortion?” We will attempt to answer this question using an EDA-informed predictive model. Given that this issue has become highly polarized by political party, we predict that liberal ideology and liberal-leaning attitudes on other issues will correlate with belief that abortion is more justified.

@@

We now extend our exploratory data analysis (EDA) to the predictor variables of interest: year of survey, ideology, Child Autonomy Index, Importance of God, Respect for Authority, and National Pride. The EDA for each variable comprises a histogram and a boxplot of the response variable, grouped by value of the predictor. We also provide a correlation matrix to detect any multicollinearity between our predictor variables, which would increase the uncertainty of our model’s parameters.

@@

*A note on data cleaning: Many survey questions were only added from 1995 onward, so observations before 1995 have a high number of missing values.  For these reasons, observations prior to 1995 were removed.

Year of survey, Respect for Authority, and National Pride were changed into factors since they are categorical variables. Year has 4 levels (one corresponding to each year), Respect for Authority has 3 levels (0 being the baseline - neutral, -1 being respect for authority bad, and 1 being respect for authority good), and National Pride is a binary (0 being the baseline - having no feelings of national pride, and 1 being having feelings of national pride) Political Ideology, Child Autonomy Index, and Importance of God had numerical sample spaces and were therefore treated as numerical variables.*

@@

$$\begin{align}y ~=~ &6.596 ~+~ 0.468x_1 ~+~ 0.33x_2 ~+~ 0.339x_3 ~+~ 0.521x_4 ~+~ 0.14x_5 \\&~-~ (0.28x_6 ~+~ 0.287x_7 ~+~ 0.179x_8 ~+~ 0.11x_9)\end{align}$$

@@

H0 (null hypothesis): An individual’s political ideology (ideology), their beliefs about child autonomy (cai), the importance of God in their life (godimportant), their view on whether respect for authority is good (respectauthority), and their pride in being American (nationalpride) are not significant predictors of aj, and the variables’ respective slope point estimates will be 0.

HA (alternative hypothesis): An individual’s political ideology (ideology), their beliefs about child autonomy (cai), the importance of God in their life (godimportant), their view on whether respect for authority is good (respectauthority), and their pride in being American (nationalpride) are significant predictors of aj, and the variables’ respective slope point estimates will not be 0.

Looking at the p-values in our output displayed above, we can reject the null hypothesis for the predictors ideology, cai, and godimportant, as the p-values are very small, way below the 0.05 significance threshold. Hence, we can conclude that the data provide convincing evidence that an individual’s political ideology, beliefs about child autonomy, and the importance of God in their life are significant predictors of their abortion attitudes.

However, we fail to reject the null hypothesis for the predictors respectauthority and nationalpride, as the p-values for each level of these predictors are above the 0.05 significance threshold. Hence, we cannot conclude that the data provide convincing evidence that an individual’s view on respect for authority and their pride in being American are significant predictors of their abortion attitudes.

@@

```
However, calculating the R-squared and root-mean-squared-error (RMSE) values of our model shows that our model is not very strong at predicting attitude toward the justifiability of abortion, with a R^2 of 0.247 and RMSE of 2.53.

@@

To see how well our model fits the data on the attitudes toward the justifiability of abortion, we conducted a 10-fold cross-validation, calculating various metrics – the Akaike information criterion (AIC), Bayesian Information Criterion (BIC) and adjusted R-squared value – to evaluate the performance of our model on 10 subsets of our data.

@@

[3 line plots, displaying AIC, BIC and R^2 value respectively, against Fold number. Each plot should have the line graph for each of the 3 models we used)

@@

Linearity [residual plot vs fitted values]

@@

Independence [residual plot vs id]

@@

[paste here after finalizing the google doc]

>>>>

Abortion is one such divisive issue in the United States. Both
pro-choice and pro-life groups have a history of mobilizing in states
across the country in support of legislation for their respective sides
(Ziegler, 2020). However, following the Supreme Court's 1973 decision in
Roe v. Wade, a ruling which protected an individual's right to have an
abortion before fetal viability, the issue has risen in political
salience. Both the pro-choice and pro-life movements have gained
national prominence, and the two major political parties have polarized
around the issue, with the Democratic Party in favor of and the
Republican Party against policies legalizing and increasing access to
abortion (Weinberger 2022). Abortion has also increasingly become a key
issue that voters consider when making their choice at the ballot box,
with an increasing share of Americans identifying as "single-issue
voters" regarding abortion (Brenan 2020).

@@

The dataset used in this analysis observes attitudes on the
justifiability of abortion among respondents in in the United States
across six "waves" of the World Values Survey (1982-2011), which is
administered every few years and collects information about people's
values and beliefs worldwide, alongside basic demographic
characteristics. The data are collected via face-to-face interviews at
the respondents' homes. In this data set, the response variable,
justifiability of abortion, is a numerical measure on a scale of 1 to 10
on the individual person's attitude toward whether abortion is
justifiable or not. Individuals responded 1 for "abortion is never
justified" and 10 for "abortion is always justified.

Our research question is as follows: "Do an individual's political
ideology and their personal attitudes/preferences towards other issues,
such as the importance of religion in their life and their respect for
authority, among others, predict their attitude on the justifiability of
abortion?" We will attempt to answer this question using an EDA-informed
predictive model. Given that this issue has become highly polarized by
political party, we predict that liberal ideology and liberal-leaning
attitudes on other issues will correlate with belief that abortion is
more justified.

@@

We now extend our exploratory data analysis (EDA) to the predictor
variables of interest: year of survey, ideology, Child Autonomy Index,
Importance of God, Respect for Authority, and National Pride. The EDA
for each variable comprises a histogram and a boxplot of the response
variable, grouped by value of the predictor. We also provide a
correlation matrix to detect any multicollinearity between our predictor
variables, which would increase the uncertainty of our model's
parameters.

@@

\*A note on data cleaning: Many survey questions were only added from
1995 onward, so observations before 1995 have a high number of missing
values. For these reasons, observations prior to 1995 were removed.

Year of survey, Respect for Authority, and National Pride were changed
into factors since they are categorical variables. Year has 4 levels
(one corresponding to each year), Respect for Authority has 3 levels (0
being the baseline - neutral, -1 being respect for authority bad, and 1
being respect for authority good), and National Pride is a binary (0
being the baseline - having no feelings of national pride, and 1 being
having feelings of national pride) Political Ideology, Child Autonomy
Index, and Importance of God had numerical sample spaces and were
therefore treated as numerical variables.\*

@@

$$\begin{aligned}y ~=~ &6.596 ~+~ 0.468x_1 ~+~ 0.33x_2 ~+~ 0.339x_3 ~+~ 0.521x_4 ~+~ 0.14x_5 \\
&~-~ (0.28x_6 ~+~ 0.287x_7 ~+~ 0.179x_8 ~+~ 0.11x_9)\end{aligned}$$

@@

H0 (null hypothesis): An individual's political ideology (ideology),
their beliefs about child autonomy (cai), the importance of God in their
life (godimportant), their view on whether respect for authority is good
(respectauthority), and their pride in being American (nationalpride)
are not significant predictors of aj, and the variables' respective
slope point estimates will be 0.

HA (alternative hypothesis): An individual's political ideology
(ideology), their beliefs about child autonomy (cai), the importance of
God in their life (godimportant), their view on whether respect for
authority is good (respectauthority), and their pride in being American
(nationalpride) are significant predictors of aj, and the variables'
respective slope point estimates will not be 0.

Looking at the p-values in our output displayed above, we can reject the
null hypothesis for the predictors ideology, cai, and godimportant, as
the p-values are very small, way below the 0.05 significance threshold.
Hence, we can conclude that the data provide convincing evidence that an
individual's political ideology, beliefs about child autonomy, and the
importance of God in their life are significant predictors of their
abortion attitudes.

However, we fail to reject the null hypothesis for the predictors
respectauthority and nationalpride, as the p-values for each level of
these predictors are above the 0.05 significance threshold. Hence, we
cannot conclude that the data provide convincing evidence that an
individual's view on respect for authority and their pride in being
American are significant predictors of their abortion attitudes.

@@

However, calculating the R-squared and root-mean-squared-error (RMSE)
values of our model shows that our model is not very strong at
predicting attitude toward the justifiability of abortion, with a R\^2
of 0.247 and RMSE of 2.53.

@@

To see how well our model fits the data on the attitudes toward the
justifiability of abortion, we conducted a 10-fold cross-validation,
calculating various metrics -- the Akaike information criterion (AIC),
Bayesian Information Criterion (BIC) and adjusted R-squared value -- to
evaluate the performance of our model on 10 subsets of our data.

@@

\[3 line plots, displaying AIC, BIC and R\^2 value respectively, against
Fold number. Each plot should have the line graph for each of the 3
models we used)

@@

Linearity \[residual plot vs fitted values\]

@@

Independence \[residual plot vs id\]

@@

\[paste here after finalizing the google doc\]
